------------------------------------------------------------------------

---
title: "Day 1"
format: pdf
editor: visual
---

## Morning: Functioning and programming in R

## w/ Dr. Cosima Meyer

Slides: <https://cosimameyer.com/wpm-summer-school/> or bit.ly/git-and-functions

and see also bit.ly/build-your-r-package and bit.ly/r-project-setup

Link to the GitHub repository: <https://github.com/cosimameyer/wpm-summer-school-r-programming>

### overviewR

A useful package that helps with checking missing values, building coverage tables and making plots

<https://github.com/cosimameyer/overviewR>

**DRY** concept = **D**on't **R**epeat **Y**ourself –\> fix it once, fix it everywhere

One function = one job. –\> a single function should do just one thing, not more. –\> if naming the function becomes complicates, then maybe split it into two or more functions

## Anatomy of a function

Name: a verb that does something

Argument: the inputs, the nouns

Body: the steps inside the curly brackets

Return statement at the end –\> not always necessary but using return() makes it more explicit

## Exercise 1:

Loading the data:

```{r}
library(readr)
ucdp_ged_sample <- read_csv("wpm-summer-school-r-programming-main/1_data/ucdp_ged_sample.csv")
```

Writing the function:

```{r}
# GOAL: write a function that takes a data frame, lowercases all of its column
# names, and replaces spaces with underscores. A genuinely useful thing when
# you've just imported a survey with columns like "Birth Year" and "Q1 Response".
#
# You have everything you need from the anatomy block:
#   name <- function(arguments) { body }

clean_colnames <- function(df) { 
  # 1. lowercase the names. make sure to assign the lower cased names back to the names of the df
  names(df) <- tolower(names(df))
  # 2. replace spaces with underscores: same here --> assign the edited column names back to the column names of the dataset
  names(df) <- gsub(" ", "_", names(df))
  # bonus task: remove the 1) and 2) from the reponses
  df[, 3] <- gsub("[0-9])", "", df[, 3])
  # 3. hand the data frame back
  return(df)
}
```

Let's see if it works...

```{r}
# --- test data (messy on purpose) -------------------------------------------
# The cells are messy too ("1) Yes"). Ignore them for now — the headers are
# the job. The cells are the bonus at the bottom of this file.
survey <- data.frame(
  check.names = FALSE,      # let R keep the ugly names as-is
  stringsAsFactors = FALSE,
  "Respondent ID" = 1:3,
  "Birth Year"    = c(1990, 1985, 2001),
  "Q1 Response"   = c("1) Yes", "2) No", "1) Yes")
)


survey                     # before
survey_clean <- clean_colnames(survey)   # after — uncomment once your function runs
survey_clean
```

The same kind of function is also part of the *janitor* package (`clean_colnames()`).

## What if the code breaks

-   Read the error message

-   `traceback()`: the trail of calls that led to the crash (what called what).

-   `print()` / `cat()` inside the function: the quick *“what’s the value right here?”* check.

-   use the `debugger`

    -   `debug(clean_colnames)` \# dig in

    -   `clean_colnames(survey)` \# pauses inside, step through it

    -   `undebug(clean_colnames)` \# climb back out

```{r}
debug(clean_colnames)         # dig in
clean_colnames(survey)        # pauses inside, step through it
undebug(clean_colnames)       # climb back out
```

Debugging some broken functions

```{r}
# Bug 1 — "nothing happened?"
clean_colnames_bug1 <- function(df) {
  names(df) <- tolower(names(df))
  names(df) <- gsub(" ", "_", names(df))
}



# Bug 2 — "it runs, but the names are still wrong"
clean_colnames_bug2 <- function(df) {
  names(df) <- tolower(names(df))
  names(df) <- gsub("_", " ", names(df))
  df
}
```

## Working with the UCDP data

```{r}
library(readr)
ged <- read_csv("wpm-summer-school-r-programming-main/1_data/ucdp_ged_sample.csv")
```

Step 1: `clean_ucdp()`

```{r}
library(dplyr)
library(janitor)
# |> is another pipe thingy, similar to %>% 
clean_ucdp <- function(dat, country_filter = NULL) {
  stopifnot(is.data.frame(dat))
  out <- dat |>
    clean_names() |>         # the janitor trick from Exercise 1
    filter(!is.na(best)) |>  # drop events with no death estimate
    rename(deaths = best)
  return(out)
}
ged <- clean_ucdp(ged)
```

Step 2: `summarize_conflicts()`

```{r}
summarize_conflict <- function(dat) {
  dat |>
    group_by(year) |>
    summarise(
      total_deaths = sum(deaths),
      n_events     = n(),
      .groups = "drop"
    )
}

summarize_conflict(ged)
```

Step 3: `plot_conflict()`

```{r}
library(ggplot2)

plot_conflict <- function(dat, title = NULL) {
  ggplot(dat, aes(x = year, y = total_deaths)) +
    geom_col(fill = "#311a36") +
    labs(title = title, x = NULL, y = "Deaths")
}
plot_conflict(ged)
```

Step 4: Bring them all together in one mega function

```{r}
run_conflict_report <- function(dat, country, title = NULL) {
  if (is.null(title)) title <- paste0("Conflict in ", country)
  dat |>
    clean_ucdp(country_filter = country) |>
    summarize_conflict() |>
    plot_conflict(title = title)
}


run_conflict_report(ged, country = "Iraq")
```

\

Exercise: extending the UCDP function to SAVE the pipeline's outputs...

continue here

```{r}
export_report <- function(dat, country, dir = "reports") {
  # 1. make sure `dir` exists (dir.create(..., recursive = TRUE))
  dir.create('C:/Users/jmstoett/OneDrive - NTNU/Documents/PhD/Conferences and workshops/2026 Methods summer school Mannheim/reports', recursive = T)
  # should be able to skip this because I work in an Rproject
  # 2. build the summary table and the plot (reuse the pipeline functions)
  dat %>% 
    clean_ucdp(country_filter = country) %>% 
    summary_table <- summarize_conflict()
  dat %>% 
    clean_ucdp(country_filter = country) %>% 
    summarize_conflict() %>% 
    summary_plot <- plot_conflict()  
  # 3. build file names from `country`
  #    hint: paste0(tolower(country), "_summary.csv")
  file_names <- paste0(tolower(country), "_summary.csv")
  # 4. write the csv (readr::write_csv) and the png (ggplot2::ggsave)
  write_csv(summary_table, "filenames")
  ggsave(filenames, plot)
  # 5. invisibly return the paths you wrote
  return()
}


export_report(ged, country = "Nigeria")
```

# Afternoon: Git & GitHub

Git is essentially version control

these commands ('git add code.R') go into the terminal

```{r}
getwd()
```

making changes
